---
title: "Cellular Impact of Contrast Agents"
author: "Angela Ariza de Schellenberger, Eric Barnhill and Shravan Vasishth"
date: "2018.05.14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#PROJ_DIR = "/home/ericbarnhill/Documents/code/R/masscyto_R"
PROJ_DIR ="./"
CYTOF_PATH <- file.path(PROJ_DIR, "cytof_data_2018_05_30.xlsx")
RTDC_PATH <- file.path(PROJ_DIR, "RT-DC_FCS_Statistics-AAS-1.xls")
knitr::opts_chunk$set(root.dir = PROJ_DIR)
setwd(PROJ_DIR)
packages <- c("gdata","tidyr","dplyr","magrittr", "ggplot2", "brms", "lattice", "HDInterval", "bayesplot")
lapply(packages, require, character.only = T)
theme_set(theme_default())
```

# Overview

This study evaluated whether novel cell measurement technologies showed differences in cell behavior between three widely used MRI contrast agents: Gadovist, Magnevist and Dotarem. These three agents are considered to have differing levels of safety and toxicity and further insight into the relative biological impact of these agents could have immediate impact for MRI protocols worldwide. To evalaute the relative toxicitiy of these agents we investigated whether we could detect differences between their effects on varying kinds of human cells. We measured differences with two different measurement techniques: Time-Of-Flight Mass Cytometry (CyTOF) and Realtime Deformation of Cells (RT-DC).

While these methods produce many thousands of output points, many of the analyses recently published with these novel technologies take point estimates on central tendency measures (such as mean or median) of sparse data sets to estimate the statistical significance of effects. These estimates are likely to be underpowered and to overstate effects, (Vashishth, Gelman et al.), fail to accurately estimate uncertainty or correlations between parameters (when they can even be estimated), and take no advantage of the distribution of thousands of data points available with each measurement. 

To optimally address these research questions, we developed an up-to-date, Bayesian statistical approach that can be applied to both of these new cell-measurement technologies as well as others. This approach incorporates quantile summary statistics to more fully estimate the distribution of individual measurements; applies orthogonal contrast codings; handles more fully specified models even with relatively sparse data, and produces a probability mass that enables posterior evaluation of the uncertainty of parameter estimates. Particularly with novel technologies and pilot studies, evaluation of uncertainty in parameter estimates is key to robust and reproducible research, and we provide a method of delivering both estimate and uncertainty that is straightforward to use, code and interpret, that may be useful for a wide range of biological measurement tools.

# CyTOF

Time-Of-Flight Mass Cytometry (CyTOF) measures spectral response of different cell types at the cellular level. Here the research question was whether there were differences in performance between three clinically used Gadolinium compounds: Gadovist, Magnevist and Dotarem. As these three compounds have different safety levels, investigating whether the signal levels were also different could aid clinical policy in choosing which compound to use.

Samples were analyzed for each of six cell types. In this pilot study a single subject was analyzed three times at three different concentration levels.

```{r load_cytof, echo=FALSE}
source('masscyto_clean_gather_data_se.R')
#mass_cyto <- read.xls(xls=CYTOF_PATH, 
#                      header=FALSE, skip=3)
data_list <- load_clean_data(CYTOF_PATH)
masscyto_data_tall <- data_list[[1]]
masscyto_data_wide <- data_list[[2]]
print(summary(masscyto_data_wide))
```

Our specific statistical question is whether there is an effect of contrast agent on the signal, independent of trial and cell type. It is clear that results will covary with cell type, but we expect the distribution around the "experiment" parameter to be iid.

## Visualizing the means

Here we plot means at concentration 1 for each contrast agent within each cell type. There seems good evidence of an effect where Dot < Mag and Mag < Gad. We use this sliding contrast coding.

```{r}
means <- subset(masscyto_data_tall, (Stat_Type == "Mean") & Contrast_Agent != "control")
means_conc_1 <- subset(means, (Concentration == 1))
means_conc_1$Contrast_Agent <- factor(means_conc_1$Contrast_Agent, c("Dotarem", "Magnevist", "Gadovist"))
plt <- ggplot(means_conc_1) + 
    geom_point(aes(x=Contrast_Agent, y=log(Value), color=Cell_Type)) +
    facet_wrap(~ Cell_Type, ncol=3) +
    ggtitle("Means By Contrast Agent @ Concentration 1") +
    scale_x_discrete(labels = c("Dot", "Mag", "Gad"))
print(plt)
```



## Log-log means plot

The log-log plot looks reasonably linear:

```{r cytof_log_log}
ggplot(subset(means, Contrast_Agent != 'control')) + 
    geom_line(aes(x=log(Concentration), y=log(Value), group=Cell_Type, color=Cell_Type)) +
    facet_wrap(~ Experiment + Contrast_Agent, scales='free')
```

Here we are interpreting the average effects. The slopes tell us the effect of GvsM and MvsD. A positive value for the respective parameter means that G > M and M > D. 


```{r cytof_brm_log_setup}

## hand-coded sliding contrasts:
masscyto_data_wide$GvsM<-ifelse(masscyto_data_wide$Contrast_Agent=="Gadovist",1,
           ifelse(masscyto_data_wide$Contrast_Agent=="Magnevist",-1,0))
masscyto_data_wide$MvsD<-ifelse(masscyto_data_wide$Contrast_Agent=="Magnevist",1,
           ifelse(masscyto_data_wide$Contrast_Agent=="Dotarem",-1,0))

priors<-c(set_prior("cauchy(0,10)", class = "b"),
          set_prior("normal(0,1)", class = "sd"), 
          set_prior("lkj(2)", class = "cor"))

brm_df <- subset(masscyto_data_wide[c("logMean", "SE", "logConcentration", "GvsM", "MvsD", "Cell_Type", "Contrast_Agent")], Contrast_Agent!="control")
brm_df$Contrast_Agent <- factor(brm_df$Contrast_Agent, labels = c("Gadovist", "Magnevist", "Dotarem"))
```

```{r cytof_brm_log,cache=TRUE}
cytof_brm_log<-brm(formula = logMean | se(SE) ~ 
            logConcentration + 
            GvsM+
            MvsD+
            logConcentration:GvsM + 
            logConcentration:MvsD +
                (1 + 
                logConcentration + 
                GvsM+
                MvsD+
                logConcentration:GvsM + 
                logConcentration:MvsD 
                | Cell_Type),
            data = brm_df, 
            family = gaussian(),
            prior = priors,
            warmup = 1000, 
            iter = 2000, 
            chains = 4,
            control = list(adapt_delta = 0.99,max_treedepth=15))

stanplot(cytof_brm_log, type="hist",pars=c("^b"))

```

Back-transforming to the scale of the dependent measure (done by Shravan):

```{r cytof_hdi_log, echo=FALSE}
Int <- unlist(posterior_samples(cytof_brm_log, pars='b_Intercept',exact_match=T))
mag_gad <- unlist(posterior_samples(cytof_brm_log, pars='b_GvsM',exact_match=T))
mag_gad_orig<-exp(Int+mag_gad)-exp(Int-mag_gad)
hist(mag_gad_orig)

dot_mag <- unlist(posterior_samples(cytof_brm_log, pars='b_MvsD',exact_match=T))
dot_mag_orig<-exp(Int+dot_mag)-exp(Int-dot_mag)
hist(dot_mag_orig)

logconc <- unlist(posterior_samples(cytof_brm_log, pars='b_logConcentration',exact_match=T))
conc_orig<-exp(Int+logconc)-exp(Int)
hist(conc_orig)

logconcGvsM <- unlist(posterior_samples(cytof_brm_log, pars='b_logConcentration:GvsM',exact_match=T))
concGvsM_orig<-exp(Int+logconcGvsM)-exp(Int-logconcGvsM)
hist(concGvsM_orig)

logconcMvsD <- unlist(posterior_samples(cytof_brm_log, pars='b_logConcentration:MvsD',exact_match=T))
concMvsD_orig<-exp(Int+logconcMvsD)-exp(Int-logconcMvsD)
hist(concMvsD_orig)

print(hdi(mag_gad_orig))
print(hdi(dot_mag_orig))
## etc.
```

How the above back-transformation works: If our model is

$log(value) = \beta_1 + \beta_2 GvsM \dots$

We take exponents on each side to get the GvsM coefficients

$exp(\beta_1 + \beta_2\times 1)-exp(\beta_1 - \beta_2\times 1)$

The above expression gives us the difference between G and M.

For the continuous measure log concentration, I back transformed as follows. let $\beta_2$ be the estimated coefficient for log concentration.

On the log scale, we have:

$log(y) = \beta_0 + \beta_2$

Taking exponents on both sides:

$y = exp(\beta_0 + \beta_2)$

Now, we can see the effect increasing concentration by 1 unit by doing:

$exp(\beta_0 + \beta_2)-exp(\beta_0)$



```{r}
mag_gad_log <-unlist(exp(posterior_samples(cytof_brm_log,'b_Intercept',exact_match=TRUE)+
                           posterior_samples(cytof_brm_log, 'b_GvsM',exact_match=TRUE)))
print('95% credible interval, log(Gad) > log(Mag) :')
print(hdi(mag_gad_log))

```

SV:
Similarly for the other case. Also, you wrote Highest (posterior) Density Interval, but this is a credible interval. The two are the same if the distribution is symmetric, but if not then they differ. The credible interval is an equal-tailed interval; the HDI not necessarily.

In the log case the relationship is multiplicative, so if 95% of the probability mass is greater than 1, there is a 95% confident effect that the contrast agent increases the signal.

SV: Is this what you mean (see code immediately below)? How can probability mass be greater than 1? I didn't understand that statement. I would just say: we can compute the posterior probability that the parameter has a positive sign. If it is positive, it means that the contrast agent G increases the signal *more* than M (the sliding contrasts do pairwise comparisons).

```{r}
mean(unlist(posterior_samples(cytof_brm_log, 'b_GvsM',exact_match=TRUE))>0)
```

## Average effects vs individual level effects for paper

### Average effects

```{r}
library(bayesplot)
post<-posterior_samples(cytof_brm_log)
## extract variances:
alpha<-post$b_Intercept
betalogconc<-post$b_logConcentration
betaGvsM<-post$b_GvsM
betaMvsD<-post$b_MvsD
betaconcGvsM<-post[,5]
betaconcMvsD<-post[,6]

cor<-posterior_samples(cytof_brm_log,"^cor")
sd<-posterior_samples(cytof_brm_log,"^sd")
sigma<-posterior_samples(cytof_brm_log,"sigma")

cell_re<-posterior_samples(cytof_brm_log,"^r_Cell")
```

```{r}
meandifflogconc<- exp(alpha + betalogconc) - exp(alpha - betalogconc)
mean(meandifflogconc)
round(quantile(meandifflogconc,prob=c(0.025,0.975)),0)

meandiffGvsM<- exp(alpha + betaGvsM) - exp(alpha - betaGvsM)
mean(meandiffGvsM)
round(quantile(meandiffGvsM,prob=c(0.025,0.975)),0)

meandiffMvsD<- exp(alpha + betaMvsD) - exp(alpha - betaMvsD)
mean(meandiffMvsD)
round(quantile(meandiffMvsD,prob=c(0.025,0.975)),0)

meandiffconcGvsM<- exp(alpha + betaconcGvsM) - exp(alpha - betaconcGvsM)
mean(meandiffconcGvsM)
round(quantile(meandiffconcGvsM,prob=c(0.025,0.975)),0)

meandiffconcMvsD<- exp(alpha + betaconcMvsD) - exp(alpha - betaconcMvsD)
mean(meandiffconcMvsD)
round(quantile(meandiffconcMvsD,prob=c(0.025,0.975)),0)

op<-par(mfrow=c(3,2),pty="s")
hist(meandifflogconc,freq=FALSE,main="Log Concentration",
     xlab="")
hist(meandiffGvsM,freq=FALSE,main="Gadovist vs Magnevist",
     xlab="")
hist(meandiffMvsD,freq=FALSE,main="Magnevist vs Dotarem",
     xlab="")
hist(meandiffconcGvsM,freq=FALSE,main="Log concentration x \n Gadovist vs Magnevist",
     xlab="")
hist(meandiffconcMvsD,freq=FALSE,main="Log concentration x \n Magnevist vs Dotarem",
     xlab="")
```

### Individual level effects

```{r}
celldifflogconc<-matrix(rep(NA,6*4000),nrow=6)
for(i in 1:6){
celldifflogconc[i,]<-exp(alpha + cell_re[,i]  + (betalogconc+cell_re[,i+6])) - 
  exp(alpha + cell_re[,i] - 
        (betalogconc+cell_re[,i+6]))
}
celldifflogconc<-t(celldifflogconc)

celldifflogconc<-as.data.frame(celldifflogconc)
colnames(celldifflogconc)<-unique(brm_df$Cell_Type)
mns <- colMeans(celldifflogconc)
celldifflogconc<-celldifflogconc[,order(mns)]
mcmc_areas(celldifflogconc)+ggtitle("Log Concentration")
```

```{r}
celldiffGvsM<-matrix(rep(NA,6*4000),nrow=6)
for(i in 1:6){
celldiffGvsM[i,]<-exp(alpha + cell_re[,i]  + (betaGvsM+cell_re[,i+7])) - 
  exp(alpha + cell_re[,i] - 
        (betaGvsM+cell_re[,i+7]))
}
celldiffGvsM<-t(celldiffGvsM)

celldiffGvsM<-as.data.frame(celldiffGvsM)
colnames(celldiffGvsM)<-unique(brm_df$Cell_Type)
mns <- colMeans(celldiffGvsM)
celldiffGvsM<-celldiffGvsM[,order(mns)]
mcmc_areas(celldiffGvsM)+ggtitle("G vs M")
```




```{r}
celldiffMvsD<-matrix(rep(NA,6*4000),nrow=6)
for(i in 1:6){
  celldiffMvsD[i,]<-exp(alpha + cell_re[,i]  + (betaMvsD+cell_re[,i+7])) - 
    exp(alpha + cell_re[,i] - 
          (betaMvsD+cell_re[,i+7]))
}
celldiffMvsD<-t(celldiffMvsD)

celldiffMvsD<-as.data.frame(celldiffMvsD)
colnames(celldiffMvsD)<-unique(brm_df$Cell_Type)
mns <- colMeans(celldiffMvsD)
celldiffMvsD<-celldiffMvsD[,order(mns)]
mcmc_areas(celldiffMvsD)+ggtitle("M vs D")
```

```{r}
celldiffconcGvsM<-matrix(rep(NA,6*4000),nrow=6)
for(i in 1:6){
  celldiffconcGvsM[i,]<-exp(alpha + cell_re[,i]  + (betaconcGvsM+cell_re[,i+7])) - 
    exp(alpha + cell_re[,i] - 
          (betaconcGvsM+cell_re[,i+7]))
}
celldiffconcGvsM<-t(celldiffconcGvsM)

celldiffconcGvsM<-as.data.frame(celldiffconcGvsM)
colnames(celldiffconcGvsM)<-unique(brm_df$Cell_Type)
mns <- colMeans(celldiffconcGvsM)
celldiffconcGvsM<-celldiffconcGvsM[,order(mns)]
mcmc_areas(celldiffconcGvsM)+ggtitle("Concentration x \n G vs M")
```

```{r}
celldiffconcMvsD<-matrix(rep(NA,6*4000),nrow=6)
for(i in 1:6){
  celldiffconcMvsD[i,]<-exp(alpha + cell_re[,i]  + (betaconcMvsD+cell_re[,i+7])) - 
    exp(alpha + cell_re[,i] - 
          (betaconcMvsD+cell_re[,i+7]))
}
celldiffconcMvsD<-t(celldiffconcMvsD)

celldiffconcMvsD<-as.data.frame(celldiffconcMvsD)
colnames(celldiffconcMvsD)<-unique(brm_df$Cell_Type)
mns <- colMeans(celldiffconcMvsD)
celldiffconcMvsD<-celldiffconcMvsD[,order(mns)]
mcmc_areas(celldiffconcMvsD)+ggtitle("Concentration x \n M vs D")
```

Could build histograms separately by cell.

```{r}
op<-par(mfrow=c(3,2),pty="s")
for(i in 1:6){
hist(celldifflogconc[,i],freq=FALSE,xlab="Log concentration",
     main=colnames(celldifflogconc)[i])
}

op<-par(mfrow=c(3,2),pty="s")
for(i in 1:6){
  hist(celldiffGvsM[,i],freq=FALSE,xlab="G vs M",
       main=colnames(celldiffGvsM)[i])
}

op<-par(mfrow=c(3,2),pty="s")
for(i in 1:6){
  hist(celldiffMvsD[,i],freq=FALSE,xlab="M vs D",
       main=colnames(celldiffMvsD)[i])
}

op<-par(mfrow=c(3,2),pty="s")
for(i in 1:6){
  hist(celldiffconcGvsM[,i],freq=FALSE,xlab="Conc x G vs M",
       main=colnames(celldiffconcGvsM)[i])
}

op<-par(mfrow=c(3,2),pty="s")
for(i in 1:6){
  hist(celldiffconcMvsD[,i],freq=FALSE,xlab="Conc x M vs D",
       main=colnames(celldiffconcMvsD)[i])
}

```



# RT-DC

RT-DC is a method of measuring cellular elastic deformations developed by TU-Dresden. We use this technology to evaluate whether cells containing common contrast agents show differences in the Young's modulus, which would indicate that their physiological properties are being altered by the contrast agent.

Distances between mean and 05 quantile are much smaller than distances between mean and 95 quantile, but in the log transform they are quite similar. So again we use a log model (in this case, there is no concentration).

```{r load_rtdc, echo=FALSE}
source('load_rtdc_4.R')

rtdc <- load_data(RTDC_PATH)
head(rtdc)
```

The RT-DC method is currently experimental, and control tests for an individual subject produce a wide range of values. Consequently each measurement with a contrast agent was paired with its own control from the same session. The Figure below shows pairwise control-treatment plots colored by subject:

```{r pairwise_plots, echo=FALSE}


rtdc_wide <- merge(subset(rtdc, Protocol == "Control"), subset(rtdc, Protocol == "Treatment"), by = c("Subject", "Date", "Cell_Type", "Contrast_Agent"), suffixes = c("_ctrl", "_tmt"))

plt <- ggplot(rtdc_wide) +
        geom_point(aes(x=E_Mean_log_ctrl,y=E_Mean_log_tmt,col=Subject)) +
        geom_abline(slope=1, alpha=0.6) + 
        facet_wrap(~ Contrast_Agent, ncol = 2) +
        ggtitle("Mean Young's Modulus")
print(plt)

```

```{r make_differences, echo=FALSE}

rtdc_diffs <- mutate(rtdc_wide, E_diff_Mean = E_Mean_log_ctrl - E_Mean_log_tmt) %>%
    mutate(., E_diff_SD = sqrt(E_SD_hi_log_ctrl^2 + E_SD_hi_log_tmt^2)) %>%
    mutate(., E_diff_SE = E_diff_SD / sqrt((Events_ctrl + Events_tmt)/2)) %>%
    select(., c("Subject", "Date", "Cell_Type", "Contrast_Agent", "E_diff_Mean", "E_diff_SE"))

print(head(rtdc_diffs, n=20))
#print(head(rtdc_wide))
```


```{r create_contrast_columns, echo=FALSE}

## Do the contrast coding manually:
rtdc_diffs$Gad_Mag<-ifelse(rtdc_diffs$Contrast_Agent=="Gadovist",-1,
                        ifelse(rtdc_diffs$Contrast_Agent=="Magnevist",1,0))
rtdc_diffs$Mag_Dot<-ifelse(rtdc_diffs$Contrast_Agent=="Magnevist",-1,
                        ifelse(rtdc_diffs$Contrast_Agent=="Dotarem",1,0))
rtdc_diffs$GdCl3_All<-ifelse(rtdc_diffs$Contrast_Agent=="GdCl3",-1,
                          ifelse(rtdc_diffs$Contrast_Agent=="Gadovist",1/3,
                                 ifelse(rtdc_diffs$Contrast_Agent=="Magnevist",1/3,
                                        ifelse(rtdc_diffs$Contrast_Agent=="Dotarem",1/3,NA)
                                 )
                          ))

diffMono<-subset(rtdc_diffs,Cell_Type=="Monocytes")
head(diffMono)
diffNeutro<-subset(rtdc_diffs,Cell_Type=="Neutrophils")
```

## Monocytes


```{r brms_monocytes,  cache=TRUE}
priors_normal_diff2 <- c(set_prior("normal(0, 1)", class = "Intercept"),
            set_prior("normal(0, 1)", class = "b"),
            set_prior("normal(0, 1)", class = "sd")
            #set_prior("normal(0, 1)", class = "sigma")
            )


mdiffMonoContrAgt_E_ME <- brm(formula = E_diff_Mean | se(E_diff_SE) ~ 1+Gad_Mag+Mag_Dot+GdCl3_All+
                (1 | Subject),
              data = diffMono, family = gaussian(), prior = priors_normal_diff2,
              iter = 4000, chains = 4, control = list(adapt_delta = 0.9999,max_treedepth=15))

stanplot(mdiffMonoContrAgt_E_ME, type="hist",pars=c("^b"))
```

Convergence diagnostics:

```{r}
summary(mdiffMonoContrAgt_E_ME,pars=c("^b","sd"))
```

```{r}
stanplot(mdiffMonoContrAgt_E_ME,type="trace",pars = c("^b","sd"))
```

The trace plots don't show any problems. I am not sure where the divergent transition is coming from. 

```{r eval=FALSE}
library(rstan)
options(mc.cores = parallel::detectCores())

diffMono$subj<-as.numeric(diffMono$Subject)
```

```{r eval=FALSE}
dat <- list(N = dim(diffMono)[1], 
            y = diffMono$E_diff_Mean,
            SD = diffMono$E_diff_SE,
            J=length(unique(diffMono$subj)),
            Gad_Mag=diffMono$Gad_Mag,
            Mag_Dot=diffMono$Mag_Dot,
            GdCl3_All=diffMono$GdCl3_All)

MonoMEfit <- stan(model_code = stanmodelcode, model_name = "MonoME", 
            data = dat, 
            iter = 2000, chains = 4) 
print(fit)
```

For Monocytes, there is weak evidence that GdCl3 causes a decrease in Young's Modulus against the other three contrast agents. The comparison of the individual contrast agents is not determinate at this time:


```{r mono_stats, echo=FALSE}
mMonoContrAgtdiffEpostME<-posterior_samples(mdiffMonoContrAgt_E_ME, "^b")
print("Prob Gad > Mag")
mean(mMonoContrAgtdiffEpostME$b_Gad_Mag>0)
print("Prob Gad < Mag")
mean(mMonoContrAgtdiffEpostME$b_Gad_Mag<0)
print("Prob Mag > Dot")
mean(mMonoContrAgtdiffEpostME$b_Mag_Dot>0)
print("Prob GdCl3 > All")
mean(mMonoContrAgtdiffEpostME$b_GdCl3_All>0)
```


## Neutrophils


```{r brms_neutrophils,cache=TRUE}
priors_cauchy_diff2 <- c(set_prior("cauchy(0, 10)", class = "Intercept"),
            set_prior("cauchy(0, 10)", class = "b"),
            set_prior("cauchy(0, 10)", class = "sd")
            )

priors_normal_diff2 <- c(set_prior("normal(0, 1)", class = "Intercept"),
            set_prior("normal(0, 1)", class = "b"),
            set_prior("normal(0, 1)", class = "sd")
            )

mdiffNeutroContrAgt_E_ME <- brm(formula = E_diff_Mean | se(E_diff_SE) ~ 1+Gad_Mag+Mag_Dot+GdCl3_All+
                (1 | Subject),
              data = diffNeutro, family = gaussian(), prior = priors_normal_diff2,
              iter = 2000, chains = 4, control = list(adapt_delta = 0.999,max_treedepth=15))

stanplot(mdiffNeutroContrAgt_E_ME, type="hist",pars=c("^b"))

mNeutroContrAgtdiffEpostME<-posterior_samples(mdiffNeutroContrAgt_E_ME, "^b")

```

For Neutrophils, there isn't much evidence for differences by Contrast Agent, though there is again maybe some weak evidence (80% likelihood) that Magnevist has higher Young's modulus than Dotarem:

```{r neutro_stats, echo=FALSE}
print("Prob Gad < Mag")
mean(mNeutroContrAgtdiffEpostME$b_Gad_Mag<0)
print("Prob Mag > Dot")
mean(mNeutroContrAgtdiffEpostME$b_Mag_Dot>0)
print("Prob GdCl3 < All")
mean(mNeutroContrAgtdiffEpostME$b_GdCl3_All<0)
```
