---
title: "Dresden analysis"
author: "Shravan Vasishth"
date: "4/30/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(brms)
theme_set(theme_default())
```

## Load and preprocess data

```{r echo=FALSE,include=FALSE}
dat<-read.table("rtdc.txt",header=TRUE)
unique(dat$Subject)

summary(dat)

## labeling is odd because each contrast agent has two labels, control and treatment, whereas control agent is by definition a single level: 
xtabs(~Contrast_Agent+Protocol,dat)

controlrows<-dat[which(dat$Protocol=="Control"),]
trtmtrows<-dat[which(dat$Protocol!="Control"),]

#trtmtrows$Contrast_Agent

summary(trtmtrows)
summary(controlrows)

controlrows$Agent<-factor("Control")
trtmtrows$Agent<-trtmtrows$Contrast_Agent

dat2<-rbind(controlrows,trtmtrows)
summary(dat2)


## relevel factors:
dat2$Agent<-factor(dat2$Agent,levels=c("Control","Dotarem","Magnevist","Gadovist","GdCl3"))
## treatment contrasts:
contrasts(dat2$Agent)

library(MASS)

## sliding contrasts:
contrasts(dat2$Agent)<-contr.sdif(5)
contrasts(dat2$Agent)

## Do the contrast coding manually:
dat2$Dot_C<-ifelse(dat2$Agent=="Control",-1,ifelse(dat2$Agent=="Dotarem",1,0))
dat2$Mag_Dot<-ifelse(dat2$Agent=="Dotarem",-1,ifelse(dat2$Agent=="Magnevist",1,0))
dat2$Gad_Mag<-ifelse(dat2$Agent=="Magnevist",-1,ifelse(dat2$Agent=="Gadovist",1,0))
dat2$GdCl3_Gad<-ifelse(dat2$Agent=="Gadovist",-1,ifelse(dat2$Agent=="GdCl3",1,0))
                                
## separate data by cell type:                            
MonoE<-subset(dat2,Cell_Type=="Monocytes" & Statistic=="Mean"  & Parameter=="E")
NeutroE<-subset(dat2,Cell_Type=="Neutrophils" & Statistic=="Mean" & Parameter=="E")

MonoA<-subset(dat2,Cell_Type=="Monocytes" & Statistic=="Mean"  & Parameter=="Area")
NeutroA<-subset(dat2,Cell_Type=="Neutrophils" & Statistic=="Mean" & Parameter=="Area")

MonoD<-subset(dat2,Cell_Type=="Monocytes" & Statistic=="Mean"  & Parameter=="Deformation")
NeutroD<-subset(dat2,Cell_Type=="Neutrophils" & Statistic=="Mean" & Parameter=="Deformation")

MonoE$Parameter<-factor(MonoE$Parameter)
NeutroE$Parameter<-factor(NeutroE$Parameter)

MonoA$Parameter<-factor(MonoA$Parameter)
NeutroA$Parameter<-factor(NeutroA$Parameter)

MonoD$Parameter<-factor(MonoE$Parameter)
NeutroD$Parameter<-factor(NeutroD$Parameter)
```

In the preprocessing (not printed out in this pdf), we separate the data from the two cell types and set up a sliding contrast coding to compare the successive effects of

- Dot(arem) vs Control 
- Mag(nevist) vs Dot
- Gad(ovist) vs Mag
- GdC13 vs Gad

A positive coefficient for a Y vs X comparison implies that the value for Y is higher than for X. 

# Area as dependent measure

## Set up priors

We will use two sets of priors: Normal(0,1) or Cauchy(0,10). The latter allows extreme values.

```{r}
priors_normal <- c(set_prior("normal(0, 10)", class = "Intercept"),
            set_prior("normal(0, 1)", class = "b"),
            set_prior("normal(0, 1)", class = "sd"),
            set_prior("normal(0, 1)", class = "sigma"),set_prior("lkj(2)", class = "cor")
            )

priors_cauchy <- c(set_prior("cauchy(0, 10)", class = "Intercept"),
            set_prior("cauchy(0, 1)", class = "b"),
            set_prior("cauchy(0, 1)", class = "sd"),
            set_prior("cauchy(0, 1)", class = "sigma"),set_prior("lkj(2)", class = "cor")
            )
```

## Using Cauchy priors

As there are more conservative, I used Cauchy priors in preference to Normal priors, but the outcome is not dependent on the priors. 

# Area

```{r cache=TRUE, include=FALSE}
mMonoA <- brm(formula = Value ~ Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad+ (1+Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad | Subject),
              data = MonoA, family = gaussian(), prior = priors_cauchy,
              iter = 2000, chains = 4, control = list(adapt_delta = 0.99))

## posteriors
mMonoApost<-posterior_samples(mMonoA, "^b")
#dim(mMonoApost)
```

Posterior probabilities of the parameter being positive:

```{r include=TRUE, print=TRUE}
(Dot_CMonoA<-round(mean(mMonoApost[,2]>0),2))
(Mag_DotMonoA<-round(mean(mMonoApost[,3]>0),2))
(Gad_MagMonoA<-round(mean(mMonoApost[,4]>0),2))
(GdCl3_GadMonoA<-round(mean(mMonoApost[,5]>0),2))
```

```{r message=FALSE,warning=FALSE}
stanplot(mMonoA, type="hist",pars=c("^b","^sd","sigma"))
```

We see some weak evidence for Dot having higher values than Control, and Mag having higher values than Dot, and Gad being higher than Mag. But GdCl3 seems to be *lower* than Gad.

```{r cache=TRUE,include=FALSE}
mNeutroA <- brm(formula = Value ~ Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad+ (1+Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad | Subject),
                data = NeutroA, family = gaussian(), prior = priors_cauchy,
                iter = 2000, chains = 4, control = list(adapt_delta = 0.99))

## posteriors
mNeutroApost<-posterior_samples(mNeutroA, "^b")
```

Posterior probabilities of the parameter being positive:

```{r include=TRUE}
(Dot_CNeutroA<-round(mean(mNeutroApost[,2]>0),2))
(Mag_DotNeutroA<-round(mean(mNeutroApost[,3]>0),2))
(Gad_MagNeutroA<-round(mean(mNeutroApost[,4]>0),2))
(GdCl3_GadNeutroA<-round(mean(mNeutroApost[,5]>0),2))
```

```{r message=FALSE,warning=FALSE}
stanplot(mNeutroA, type="hist",pars=c("^b","^sd","sigma"))
```

Very similar results from Neurophils.


# Deformation

```{r cache=TRUE, include=FALSE}
mMonoD <- brm(formula = Value ~ Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad+ (1+Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad | Subject),
              data = MonoD, family = gaussian(), prior = priors_cauchy,
              iter = 2000, chains = 4, control = list(adapt_delta = 0.99))

## posteriors
mMonoDpost<-posterior_samples(mMonoD, "^b")
#dim(mMonoDpost)
```

Posterior probabilities of the parameter being positive:

```{r include=TRUE, print=TRUE}
(Dot_CMonoD<-round(mean(mMonoDpost[,2]>0),2))
(Mag_DotMonoD<-round(mean(mMonoDpost[,3]>0),2))
(Gad_MagMonoD<-round(mean(mMonoDpost[,4]>0),2))
(GdCl3_GadMonoD<-round(mean(mMonoDpost[,5]>0),2))
```

```{r message=FALSE,warning=FALSE}
stanplot(mMonoD, type="hist",pars=c("^b","^sd","sigma"))
```

[Analysis of deformation to come]

```{r cache=TRUE,include=FALSE}
mNeutroD <- brm(formula = Value ~ Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad+ (1+Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad | Subject),
                data = NeutroD, family = gaussian(), prior = priors_cauchy,
                iter = 2000, chains = 4, control = list(adapt_delta = 0.99))

## posteriors
mNeutroDpost<-posterior_samples(mNeutroD, "^b")
```

Posterior probabilities of the parameter being positive:

```{r include=TRUE}
(Dot_CNeutroD<-round(mean(mNeutroDpost[,2]>0),2))
(Mag_DotNeutroD<-round(mean(mNeutroDpost[,3]>0),2))
(Gad_MagNeutroD<-round(mean(mNeutroDpost[,4]>0),2))
(GdCl3_GadNeutroD<-round(mean(mNeutroDpost[,5]>0),2))
```

```{r message=FALSE,warning=FALSE}
stanplot(mNeutroD, type="hist",pars=c("^b","^sd","sigma"))
```


# Young's Modulus

```{r cache=TRUE, include=FALSE}
mMonoE <- brm(formula = Value ~ Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad+ (1+Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad | Subject),
              data = MonoE, family = gaussian(), prior = priors_cauchy,
              iter = 2000, chains = 4, control = list(adapt_delta = 0.99))

## posteriors
mMonoEpost<-posterior_samples(mMonoE, "^b")
#dim(mMonoEpost)
```

Posterior probabilities of the parameter being positive:

```{r include=TRUE, print=TRUE}
(Dot_CMonoE<-round(mean(mMonoEpost[,2]>0),2))
(Mag_DotMonoE<-round(mean(mMonoEpost[,3]>0),2))
(Gad_MagMonoE<-round(mean(mMonoEpost[,4]>0),2))
(GdCl3_GadMonoE<-round(mean(mMonoEpost[,5]>0),2))
```

```{r message=FALSE,warning=FALSE}
stanplot(mMonoE, type="hist",pars=c("^b","^sd","sigma"))
```

[Analysis of deformation to come]

```{r cache=TRUE,include=FALSE}
mNeutroE <- brm(formula = Value ~ Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad+ (1+Dot_C+Mag_Dot +Gad_Mag+GdCl3_Gad | Subject),
                data = NeutroE, family = gaussian(), prior = priors_cauchy,
                iter = 2000, chains = 4, control = list(adapt_delta = 0.99))

## posteriors
mNeutroEpost<-posterior_samples(mNeutroE, "^b")
```

Posterior probabilities of the parameter being positive:

```{r include=TRUE}
(Dot_CNeutroE<-round(mean(mNeutroEpost[,2]>0),2))
(Mag_DotNeutroE<-round(mean(mNeutroEpost[,3]>0),2))
(Gad_MagNeutroE<-round(mean(mNeutroEpost[,4]>0),2))
(GdCl3_GadNeutroE<-round(mean(mNeutroEpost[,5]>0),2))
```

```{r message=FALSE,warning=FALSE}
stanplot(mNeutroE, type="hist",pars=c("^b","^sd","sigma"))
```


## Conclusion

# Area

Monocytes: The probability (given data and model) that

-  Dot has a higher value than control:  `r Dot_CMonoA`
- Mag has a higher value than Dot: `r Mag_DotMonoA`
- Gad has a higher value than Mag: `r Gad_MagMonoA`
- GdCl3 has a higher value than Gad: `r GdCl3_GadMonoA`

Neutrophils: The probability (given data and model) that

-  Dot has a higher value than control:  `r Dot_CNeutroA`
- Mag has a higher value than Dot: `r Mag_DotNeutroA`
- Gad has a higher value than Mag: `r Gad_MagNeutroA`
- GdCl3 has a higher value than Gad: `r GdCl3_GadNeutroA`

# Deformation

Monocytes: The probability (given data and model) that

-  Dot has a higher value than control:  `r Dot_CMonoD`
- Mag has a higher value than Dot: `r Mag_DotMonoD`
- Gad has a higher value than Mag: `r Gad_MagMonoD`
- GdCl3 has a higher value than Gad: `r GdCl3_GadMonoD`

Neutrophils: The probability (given data and model) that

-  Dot has a higher value than control:  `r Dot_CNeutroD`
- Mag has a higher value than Dot: `r Mag_DotNeutroD`
- Gad has a higher value than Mag: `r Gad_MagNeutroD`
- GdCl3 has a higher value than Gad: `r GdCl3_GadNeutroD`

# Young's Modulus

Monocytes: The probability (given data and model) that

-  Dot has a higher value than control:  `r Dot_CMonoE`
- Mag has a higher value than Dot: `r Mag_DotMonoE`
- Gad has a higher value than Mag: `r Gad_MagMonoE`
- GdCl3 has a higher value than Gad: `r GdCl3_GadMonoE`

Neutrophils: The probability (given data and model) that

-  Dot has a higher value than control:  `r Dot_CNeutroE`
- Mag has a higher value than Dot: `r Mag_DotNeutroE`
- Gad has a higher value than Mag: `r Gad_MagNeutroE`
- GdCl3 has a higher value than Gad: `r GdCl3_GadNeutroE`

So, I would conclude that given that we have only 5 people's data, there might be grounds for concluding that we can use different contrast agents on Monocytes and Neutrophils to study the effect of the agents on the cells. Only the dependent measure Area shows the expected increase by contrast agent (ordered from least toxic to most toxic). The only thing that may be odd is that the most toxic control, GdCl3, does not have higher values than Gad. If anything, Gad seems to have higher values; so Gad seems to be even more toxic than GdC13? Not sure what to make of that.

# Analysis of Control vs Treatment

```{r}
load("df_differences.RData")
#head(rtdc_diff)
diffdat<-subset(rtdc_diff,Parameter=="E" & Statistic=="Mean")
diffMono<-subset(diffdat,Cell_Type=="Monocytes")
diffNeutro<-subset(diffdat,Cell_Type=="Neutrophils")
#summary(diffMono)
#summary(diffNeutro)
```

```{r}
library(lattice)
xyplot(diff~Subject|Cell_Type,diffdat)
```

Set priors (we will use cauchy priors):

```{r}
priors_normal_diff <- c(set_prior("normal(0, 10)", class = "Intercept"),
            set_prior("normal(0, 1)", class = "sd"),
            set_prior("normal(0, 1)", class = "sigma"))

priors_cauchy_diff <- c(set_prior("cauchy(0, 10)", class = "Intercept"),
            set_prior("cauchy(0, 10)", class = "sd"),
            set_prior("cauchy(0, 10)", class = "sigma"))
```

```{r cache=TRUE}
mdiffMonoE <- brm(formula = diff ~ 1+(1 | Subject),
              data = diffMono, family = gaussian(), prior = priors_cauchy_diff,
              iter = 2000, chains = 4, control = list(adapt_delta = 0.999))

mdiffNeutroE <- brm(formula = diff ~ 1+(1 | Subject),
              data = diffMono, family = gaussian(), prior = priors_cauchy_diff,
              iter = 2000, chains = 4, control = list(adapt_delta = 0.999))

```

The posteriors don't show convincing evidence for any difference in either cell:

```{r}
## posteriors
mdiffMonoEpost<-posterior_samples(mdiffMonoE, "Intercept")
mean(mdiffMonoEpost>0)

## posteriors
mdiffNeutroEpost<-posterior_samples(mdiffNeutroE, "Intercept")
mean(mdiffNeutroEpost>0)
```

Graphical summaries:
```{r message=FALSE,warning=FALSE}
stanplot(mdiffMonoE, type="hist",pars=c("Intercept","^sd","sigma"))

stanplot(mdiffNeutroE, type="hist",pars=c("Intercept","^sd","sigma"))
```

## Combined cells

Maybe it makes sense to have both cells' data combined and just control for Cell differences:

```{r}
priors_normal_diff_cell <- c(set_prior("normal(0, 10)", class = "Intercept"),
            set_prior("normal(0, 1)", class = "b"),
            set_prior("normal(0, 1)", class = "sd"),
            set_prior("normal(0, 1)", class = "sigma"),set_prior("lkj(2)", class = "cor")
            )

priors_cauchy_diff_cell <- c(set_prior("cauchy(0, 10)", class = "Intercept"),
            set_prior("cauchy(0, 10)", class = "b"),
            set_prior("cauchy(0, 10)", class = "sd"),
            set_prior("cauchy(0, 10)", class = "sigma"),set_prior("lkj(2)", class = "cor")
            )
```

```{r}
mdiffE <- brm(formula = diff ~ 1+Cell_Type+(1+Cell_Type | Subject),
              data = diffdat, family = gaussian(), prior = priors_cauchy_diff_cell,
              iter = 2000, chains = 4, control = list(adapt_delta = 0.999))

stanplot(mdiffE, type="trace",pars=c("Intercept","^b","^sd","sigma"))

stanplot(mdiffE, type="hist",
         pars=c("Intercept","^b","^sd","sigma"))
```

In Young's modulus, I don't see any convincing evidence for an effect in either the overall effect (the intercept), or in individual subjects' effects, except Philip, whose intercept seems to be positive; but there isn't a lot of data from Philip; Gela has the most data and he/she shows no effects.

# Looking at differences by Contrast Agent

Angela wrote: 

- The main variable to be compared is the Young´s modulus (E).

Eric also wrote:

- we were interested in a further analysis that had no control group by itself, but rather four groups, one for each contrast agent (Gad, Dot, Mag, GdCl3), whose measurement points are the difference between each subject's control and treatment measurements for that contrast agent. So, the measurement point for Angela in the Gad group is (Angela-Gad-treatment minus Angela-Gad-control)  and the measurement point for Markus in Dotarem is (Markus-Dot-treatment minus Markus-Dot-control). 

I implement this analysis next:

```{r}
diffdat<-subset(rtdc_diff,Parameter=="E" & Statistic=="Mean")

## Do the contrast coding manually:
diffdat$Mag_Dot<-ifelse(diffdat$Contrast_Agent=="Dotarem",-1,
                        ifelse(diffdat$Contrast_Agent=="Magnevist",1,0))
diffdat$Gad_Mag<-ifelse(diffdat$Contrast_Agent=="Magnevist",-1,
                        ifelse(diffdat$Contrast_Agent=="Gadovist",1,0))
diffdat$GdCl3_Gad<-ifelse(diffdat$Contrast_Agent=="Gadovist",-1,
                          ifelse(diffdat$Contrast_Agent=="GdCl3",1,0))

diffMono<-subset(diffdat,Cell_Type=="Monocytes")
diffNeutro<-subset(diffdat,Cell_Type=="Neutrophils")
```

## Monocytes

```{r}
priors_cauchy_diff2 <- c(set_prior("cauchy(0, 10)", class = "Intercept"),
            set_prior("cauchy(0, 10)", class = "b"),
            set_prior("cauchy(0, 10)", class = "sd"),
            set_prior("cauchy(0, 10)", class = "sigma")
            )


mdiffMonoContrAgt_E <- brm(formula = diff ~ 1+Mag_Dot+Gad_Mag+GdCl3_Gad+
                (1| Subject),
              data = diffMono, family = gaussian(), prior = priors_cauchy_diff2,
              iter = 2000, chains = 4, control = list(adapt_delta = 0.999))

stanplot(mdiffMonoContrAgt_E, type="trace",pars=c("Intercept","^b","^sd","sigma"))

stanplot(mdiffMonoContrAgt_E, type="hist",pars=c("Intercept","^b","^sd","sigma"))

mMonoContrAgtdiffEpost<-posterior_samples(mdiffMonoContrAgt_E, "^b")
```

For Monocytes, one could argue that there is some *weak* evidence for increasing differences by Contrast Agent:


```{r}
mean(mMonoContrAgtdiffEpost$b_Mag_Dot>0)
mean(mMonoContrAgtdiffEpost$b_Gad_Mag>0)
mean(mMonoContrAgtdiffEpost$b_GdCl3_Gad>0)
```

## Neutrophils

```{r}
mdiffNeutroContrAgt_E <- brm(formula = diff ~ 1+Mag_Dot+Gad_Mag+GdCl3_Gad+
                (1| Subject),
              data = diffNeutro, family = gaussian(), prior = priors_cauchy_diff2,
              iter = 2000, chains = 4, control = list(adapt_delta = 0.999))

stanplot(mdiffNeutroContrAgt_E, type="trace",pars=c("Intercept","^b","^sd","sigma"))

stanplot(mdiffNeutroContrAgt_E, type="hist",pars=c("Intercept","^b","^sd","sigma"))

mNeutroContrAgtdiffEpost<-posterior_samples(mdiffNeutroContrAgt_E, "^b")
```

For Neutrophils, there isn't much evidence for increasing differences by Contrast Agent:

```{r}
mean(mNeutroContrAgtdiffEpost$b_Mag_Dot>0)
mean(mNeutroContrAgtdiffEpost$b_Gad_Mag>0)
mean(mNeutroContrAgtdiffEpost$b_GdCl3_Gad>0)
```