---
title: "RT-DC Data"
output:
  pdf_document: default
  html_notebook: default
---

Setup

```{r setup, message=FALSE, error=FALSE, results=FALSE, warning=FALSE}
PROJ_DIR = "/home/ericbarnhill/Documents/code/R/masscyto_R"
knitr::opts_chunk$set(root.dir = PROJ_DIR)
setwd(PROJ_DIR)
packages <- c("gdata","tidyr","dplyr","magrittr", "ggplot2")
lapply(packages, require, character.only = T)
```

Load and clean data
```{r}
source("load_rtdc.R")
XL_PATH <- file.path(PROJ_DIR, "rt-dc_eb.xls")
rtdc_tall <- load_rtdc(XL_PATH)

```
Explore summary stats of Young's modulus and deformation. There are many outliers so the effect is clearest using a robust statistic (median):

```{r}
    FUN = median
    LOG = F;
    plt_e <- rtdc_exploratory_boxplot(rtdc_tall, "mean_e", "Mean Young's Modulus", FUN, LOG)
    print(plt_e)
    plt_def <- rtdc_exploratory_boxplot(rtdc_tall, "mean_def", "Mean Deformation", FUN, LOG)
    print(plt_def)
```

Log results look very similar, also the data is not probably not a good fit for the log: deformation has a range of (0,1) and Young's modulus can be negative.

To get a sense of the distributions, look at histograms of controls (excluding GdCl3 Neutrophils Young's which seems to be off somehow) for one cell type

```{r}
controls_e <-rtdc_tall %>%
    subset(prot == "Control" & measurement == "mean_e" & cell_type == "Neutrophils" & !(cell_type == "Neutrophils" & cont_ag == "GdCl3"))
controls_def <- rtdc_tall %>%
    subset(prot == "Control" & measurement == "mean_def" & cell_type == "Neutrophils")
plt_e <- ggplot(controls_e) + geom_histogram(aes(x=value), bins=10) + ggtitle("Histogram for E Controls")
plt_def <- ggplot(controls_def) + geom_histogram(aes(x=value), bins=10) + ggtitle("Histogram for Deformation Controls")
print(plt_e)
print(plt_def)
```

Both E and deformation have outliers. Looking back at the strip plot, each measurement appears to have one heavy outlier. The E is reasonably normal. The deformation looks sparse and this needs to be understood.